---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
data <- read.csv("activity.csv", header = TRUE)
str(data)
```


## What is mean total number of steps taken per day?

Calculate the sum of steps for each day:
```{r}
stepsPerDay <- tapply(data$steps, data$date, sum, na.rm =TRUE)
```

Plot a histogram of this data:

```{r}
hist(stepsPerDay)
```

###Mean
To calculate the mean of the data, we simply run:
```{r}
meanSteps <- mean(stepsPerDay)
```
and see that the mean is `r meanSteps`.  

###Median
The median is calculated simillarly by running:
```{r}
medianSteps <- median(stepsPerDay)
```

Which shows that the median is `r medianSteps`.

## What is the average daily activity pattern?
First, we create an array of the averages of each 5-minute interval
over all the obersvations.
We then plot this data to get an overview.

```{r}
intervalAvgs <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
plot(names(intervalAvgs), intervalAvgs, type="l")
```

We can see from the plot that the maximum value is somewhere between 0700 and 1000.
To get the exact result, we can run

```{r}
which.max(intervalAvgs)
```
and see that the interval "835", which is the 104th element in the *internalAvgs* array,
contains the maximum number of steps.
To get the actual number, we simply acces this element:
```{r}
intervalAvgs[104]
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
